---
title: "Report"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

```{r cars}

```

## Explanatory data analysis

We first examine the descriptive statistics of the response and all predictors.

```{r indiv_boxplots, echo=FALSE}
covid <- read.table("pblc_bhv_covid.csv",sep = ',', header=TRUE)
boxplot(covid[2:6], xlab="Variables", ylab="Values", main="Boxplots of variables")
boxplot(covid[6:10], xlab="Variables", ylab="Values", main="Boxplots of variables")
```
It is clear that all variables have very different scales from each other, so models with unstandardized and standardized variables should be examined.

We plot all predictors against the number of cases.

```{r plots_against_cases, echo=FALSE}

par(mfrow=c(3,3))

with(covid,plot(distancing,cases))
with(covid,plot(bar_visit,cases))
with(covid,plot(large_events,cases))
with(covid,plot(mask_prop,cases))
with(covid,plot(other_mask_prop,cases))
with(covid,plot(public_transit,cases))
with(covid,plot(resto_visit,cases))
with(covid,plot(worked_outside,cases))
```

`distancing`, `bar_visit` and `large_events` appear to have two clusters with different mean case counts. On the other hand, `mask_prop`, `other_mask_prop`, `resto_visit` and `worked_outside` appear to have two trend lines. `public_transit` also looks like there are two trend lines, but both are quite flat. This suggests that modelling `distancing`, `bar_visit` and `large_events` as two-level factors with interactions with the remaining predictors might be profitable. A tree-based regression method might also be worth exploring.

We will now examine the covariance matrix of the predictors.

```{r}
predictors <- subset(covid, select=-c(cases,date))

cor_pred <- cor(as.matrix(predictors))

for (i in 1:ncol(predictors)) {
  print(cor_pred[i, abs(cor_pred[i,]) > 0.6])
}
```

There is some substantial correlation between some predictors, which can be further explored with plots. In particular, `bar_visit`, `mask_prop`, `large_events` and `other_mask_prop` seem correlated with each other. This suggests that a regularization approach can be profitable.